Plastic in the Urban Freshwater Environment

### 2021-10-28 analysis.Rmd
various edits
update exploration figures, manuscript additions, adding data index and populating readme files

>>>>

#load packages and data
```{r}
library(ggplot2) #for plotting
library(broom) #for cleaning up output from lm()
library(here) #for data loading/saving
library(tidymodels) #for modeling

data <- readRDS(here("data", "processed_data", "mpwrfcombo.rds"))
```

# fit linear model to continuous outcome (particles_l)
```{r}
#attempt to visualize data
#look at particles per L vs population
ggplot(data, aes(x = population, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Particles/Liter")
```

# particles per liter vs population with outliers filtered out
```{r}
#attempt to visualize data
#look at particles per L vs population
data %>% filter(particles_l < 750) %>% ggplot(aes(x = population, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Particles/Liter")
```

# fit linear model to continuous outcome (particles_l) with log transformation
```{r}
#attempt to visualize data
#look at particles per L vs population
ggplot(data, aes(x = population, y = particles_l)) + geom_point() +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Log(Particles/Liter)")
```

# visualize particles vs e.coli.cfu 
```{r}
ggplot(data, aes(x = e.coli.cfu, y = particles_l)) + geom_point() + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. E. coli CFU", x = "CFU/mL", y = "Particles/Liter")
```

# particles per liter vs e.coli.cfu with log transformation
```{r}
ggplot(data, aes(x = e.coli.cfu, y = particles_l)) + geom_point() + scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + geom_smooth(method = lm) + labs(title = "Microplastic Concentration vs. E. coli CFU", x = "Log(CFU/mL)", y = "Log(Particles/Liter)")
```

# linear regression
```{r}
#linear regression model specification set up
lm_mod <- linear_reg() %>% set_engine("lm")
#estimating/training linear model
lm_fit1 <- lm_mod %>% fit(particles_l ~ ., data = data)
#view summary of fit
tidy(lm_fit1)
```

# linear regression for one predictor, one outcome
```{r}
#predict particles per liter based on population
lm_fit2 <- lm_mod %>% fit(particles_l ~ population, data = data)
#view summary of fit
tidy(lm_fit2)

#predict particles per liter based on e coli
lm_fit3 <- lm_mod %>% fit(particles_l ~ e.coli.cfu, data = data)
#view summary of fit
tidy(lm_fit3)

#comparison of fit
comp2_3 <- anova(lm_fit2$fit, lm_fit3$fit, test = "Chisq")
comp2_3
```

Here, we get an error message that models were not fitted to the same size of dataset. This is likely due to missing values in some predictors. To solve this, we can clean up the dataset some more and create a dataset specifically to use for modelling, that has only predictors of interest and no missing values.

#clean up data for modelling
```{r}
data %>% is.na() %>% summary()
complete <- data %>% na.omit()
```

Trying code from above again with the complete dataset

```{r}
#estimating/training linear model
lm_fit4 <- lm_mod %>% fit(particles_l ~ ., data = complete)
#view summary of fit
tidy(lm_fit4)
#predict particles per liter based on population
lm_fit5 <- lm_mod %>% fit(particles_l ~ population, data = complete)
#view summary of fit
tidy(lm_fit5)

#predict particles per liter based on e coli
lm_fit6 <- lm_mod %>% fit(particles_l ~ e.coli.cfu, data = complete)
#view summary of fit
tidy(lm_fit6)

#comparison of fit
comp5_6 <- anova(lm_fit5$fit, lm_fit6$fit, test = "Chisq")
comp5_6
```

Here, we have a VERY large RSS, which I think indicates a poor model fit for our data. Let's try with some variables removed.

# Subsetting data to remove unimportant variables
```{r}
subset <- complete %>% select(-"site", 
                              "lat", 
                              "long",
                              "date") 
#estimating/training linear model
lm_fit7 <- lm_mod %>% fit(particles_l ~ ., data = subset)
#view summary of fit
tidy(lm_fit7)

#predict particles per liter based on population
lm_fit8 <- lm_mod %>% fit(particles_l ~ population, data = subset)
#view summary of fit
tidy(lm_fit8)

#predict particles per liter based on e coli
lm_fit9 <- lm_mod %>% fit(particles_l ~ e.coli.cfu, data = subset)
#view summary of fit
tidy(lm_fit9)

#model comparison
comp8_9 <- anova(lm_fit8$fit, lm_fit9$fit, test = "Chisq")
comp8_9
```

Here, we see the same RSS results even after subsetting the data (the values are exactly the same when run on the "complete" data as on the "subset" data.)

# model evaluation, splitting data into train and test
```{r}
#split into test and train
#set seed for reproducible analysis
set.seed(222)
#subset 3/4 of data as training set
data_split <- initial_split(data, prop = 3/4)

#save sets as data frames
train_data <- training(data_split)
test_data <- testing(data_split)
```

# recipe to fit continuous outcome to all predictors
```{r}
#recipe for particles per liter, all predictors
mp_rec <- 
  recipe(particles_l ~ ., data = train_data)

#linear regression model workflow set up
mp_wflow <- 
  workflow() %>% add_model(lm_mod) %>% add_recipe(mp_rec)

#use workflow to prepare recipe and train model with predictors
mp_fit <- 
  mp_wflow %>% fit(data = train_data)

#extract model coefficient
mp_fit %>% extract_fit_parsnip() %>% tidy()
```

Here, we see site names come up first, even though we know site names are not going to be good predictors of microplastic levels. I need to go back and remove the variables that aren't going to be helpful predictors (ex: site name)

# evaluate model with RMSE
```{r}
#use trained workflow to predict with test data
predict(mp_fit, test_data)

#include probabilities
mp_aug <- augment(mp_fit, test_data)

#get RMSE for test data
mp_aug %>% rmse(truth = particles_l, .pred)

```

Why are there so many rows resulting? When we did the analysis last week, only one row was the result, and it was just rmse and provided a value around 1.

#trying model evaluation with SUBSET
```{r}

```

### 2021-10-28 code/analysis_code/exploration.Rmd
various edits
update exploration figures, manuscript additions, adding data index and populating readme files

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)

#path to data
data_location <- here::here("data","processed_data","popdata.rds")

@@

# take a look at data
```{r}
summary(mpdata)

@@

p1 <- mpdata %>% ggplot(aes(x = date, y = particles_l)) + geom_boxplot(aes()) + labs(title = "Microplastic Concentration")

@@

#save figure
figure_file = here("results","concentrationbydate.png")
ggsave(filename = figure_file, plot=p1)
```

@@

# MP level vs population
```{r}
p2 <- mpdata %>% ggplot(aes(x = population, y = particles_l)) + geom_point()

@@

#save figure
figure_file = here("results","particlesvpop.png")
ggsave(filename = figure_file, plot=p1)
```

@@

```{r}
mpdata %>% ggplot(aes(x = e.coli.cfu, y = particles_l)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust=1))
```

Something weird is giong on with the values for e.coli.cfu.

```{r}
class(mpdata$e.coli.cfu)
```

It looks like that ended up saved as a character instead of integer. Let's convert. 
# covert character to numeric
```{r}
mpdata <- mpdata %>% mutate(e.coli.cfu = as.numeric(e.coli.cfu))
class(mpdata$e.coli.cfu)
```

# try plot again
```{r}
p3 <- mpdata %>% ggplot(aes(x = e.coli.cfu, y = particles_l)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust=1))

plot(p3)

#save figure
figure_file = here("results","mpvbacteria.png")
ggsave(filename = figure_file, plot=p3)
```

@@

```{r}
p4 <- mpdata %>% ggplot(aes(x = e.coli.cfu, y = particles_l)) + geom_point() + theme(axis.text.x = element_text(angle = 90, hjust=1)) + scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10')

#save figure
figure_file = here("results","logmpvbacteria.png")
ggsave(filename = figure_file, plot=p4)
```

Note: missing values are for e.coli.cfu for July 2021 sampling. These values should be available for comparison soon.

Let's look at overall distribution of MP levels
```{r}
mpdata %>% ggplot(aes(x = particles_l)) + geom_histogram()
```

@@

```{r}
p5 <- mpdata %>% filter(particles_l < 400) %>% ggplot(aes(x = particles_l)) + geom_histogram()

#save figure
figure_file = here("results","distribution.png")
ggsave(filename = figure_file, plot=p5)
```

We see that most levels fall below 100 particles per liter.

```{r}
summary(mpdata$particles_l)

@@

I also wanted to look at proximity of sample sites to water reclamation facilities (WRFs). Let's try that. 
# load WRF location
```{r}
#path to data
wrfdata_location <- here::here("data","processed_data","wrflocation.rds")

#load data
wrflocation <- readRDS(wrfdata_location)
```

Eventually, I would like to be able to view this information on a map - where WRFs are labelled, and samples sites are color-coded by density of microplastics found. For now, let's see if we can calculate distance otherwise.

```{r}
library(geodist)

#calculate distance between sample coordinates and WRF location
wrfdistance <- geodist_vec(mpdata$long, mpdata$lat, wrflocation$long, wrflocation$lat)
```

I think the product we have here is a matrix of the distance of each sample site from each of the three WRFs, in the order that they appear in the WRF data (Cedar Creek, North Oconee, Middle Oconee). But, the labelling of the data is not super clear, so I need to verify this to be sure.

```{r}
test <- mpdata %>% mutate(wrfdist = geodist_vec(mpdata$long, mpdata$lat, wrflocation$long, wrflocation$lat))
```

This produced another column like I wanted, but I don't know which WRF the distance was calculated for. Maybe I need to save WRF locations as individual lat and long values, and produce three new columns that way? However, I am not sure what the final product would look like if we do that, because a sample site could be closely associated with one WRF but much further from another one. Maybe I need to narrow this down to proximity to the nearest WRF, so that each sample site is only associated with the distance to ONE WRF, not three.

>>>>

# load packages and data
```{r}
library(dplyr) #for data processing
library(here) #to set paths
library(tidyverse)
library(ggplot2)
library(geodist) #to calculate distance from sample site to each water reclamation facility

#path to data
data_location <- here::here("data","processed_data","mpwrfcombo.rds")

@@

# MP concentration by season
```{r}
summary(mpdata)

@@

mpdata$date <- factor(mpdata$date , levels=c("nov_20", "feb_21", "apr_21", "july_21"))

p1 <- mpdata %>% ggplot(aes(x = date, y = particles_l)) + 
  geom_boxplot(aes()) + 
  labs(title = "Microplastic Concentration by Sample Date", x = "Date", y = "Particles/Liter") + 
  scale_x_discrete(labels=c("Fall 2020",
                            "Winter 2021",
                            "Spring 2021",
                            "Summer 2021"))

@@

#save figure
ggsave(filename = here("results","concentrationbydate.png"), plot=p1)
```

@@

# MP level vs population
```{r}
p2 <- mpdata %>% ggplot(aes(x = population, y = particles_l)) + geom_point() +
  labs(title = "Microplastic Concentration vs Population", x = "Population", y = "Particles/Liter")

@@

#save figure
ggsave(filename = here("results","particlesvpop.png"), plot=p2)
```

@@

# particles vs population with log transformation
```{r}
p3 <- mpdata %>% ggplot(aes(x = population, y = particles_l)) + 
  geom_point() + 
  scale_y_continuous(trans='log10') +
  labs(title = "Microplastic Concentration vs. Population", x = "Population", y = "Log(Particles/Liter)")

plot(p3)
#save figure
figure_file = here("results","logparticlesvpop.png")
ggsave(filename = figure_file, plot=p3)
```

# plot e.coli.cfu vs particles_l
```{r}
p4 <- mpdata %>% ggplot(aes(x = e.coli.cfu, y = particles_l)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) +
  labs(title = "Microplastic Concentration vs. E. coli CFU", x = "CFU", y = "Particles/Liter")

#save figure
figure_file = here("results","mpvcfu.png")
ggsave(filename = figure_file, plot=p4)
```

@@

```{r}
p5 <- mpdata %>% ggplot(aes(x = e.coli.cfu, y = particles_l)) + 
  geom_point() + 
  theme(axis.text.x = element_text(angle = 90, hjust=1)) + 
  scale_x_continuous(trans='log10') +
  scale_y_continuous(trans='log10') + 
  labs(title = "Microplastic Concentration vs. E. coli CFU", x = "Log(CFU)", y = "Log(Particles/Liter)")

#save figure
figure_file = here("results","logmpvcfu.png")
ggsave(filename = figure_file, plot=p5)
```

Let's look at overall distribution of MP levels.

# Distribution of MP concentration
```{r}
mpdata %>% ggplot(aes(x = particles_l)) + geom_histogram() + 
  labs(title = "Distribution of Microplastic Concentration", x = "Particles/Liter", y = "Observations")
```

@@

# Distribution of MP concentration without outliers
```{r}
p6 <- mpdata %>% filter(particles_l < 400) %>% ggplot(aes(x = particles_l)) + geom_histogram() + 
  labs(title = "Distribution of Microplastic Concentration", x = "Particles/Liter", y = "Observations")

plot(p6)

#save figure
figure_file = here("results","distribution.png")
ggsave(filename = figure_file, plot=p6)
```

We see that most levels fall below 100 particles per liter. 
# Summary statistics
```{r}
summary(mpdata$particles_l)
```

@@

# correlation matrices
```{r}
library(GGally) #for produces correlation matrices

#subsetting data to keep only variables we are interested in for the corr plot
mpdatacorr <- mpdata %>% select(-"quarter", 
                                -"month", 
                                -"lat",
                                -"long", 
                                -"year")

#correlation using MP, UOWN, population data
p7 <- ggcorr(mpdatacorr, 
       label_alpha = TRUE) + labs(title = "Microplastic Predictor Correlation Matrix")

plot(p7)
#save figure
figure_file = here("results","matrix.png")
ggsave(filename = figure_file, plot=p7)
```

# exploring land use with a correlation matrix
```{r}
athensland <- readRDS(here("data", "processed_data", "athensland.rds"))

#subset data to view in plot
athenslandcorr <- athensland %>% select(-"long", 
                                        -"lat")

#land use correlation matrix
p8 <- ggcorr(athenslandcorr, 
       label_alpha = TRUE) + labs(title = "Land Use Correlation Matrix")

plot(p8)
#save figure
figure_file = here("results","landmatrix.png")
ggsave(filename = figure_file, plot=p8)
```

### 2021-10-28 products/manuscript/Manuscript.Rmd
various edits
update exploration figures, manuscript additions, adding data index and populating readme files

---
title: "Microplastics in the Urban Environment"
subtitle: "Exploratory Study of Freshwater Microplastics in Athens, GA"
author: Amelia M. Foley
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
  bookdown::html_document2: 
    toc: false
bibliography: library.bib #I am having trouble with the bib file (error message: pandoc error 83 file not found). Removing this line for now so that Manuscript will knit. Will continue to troubleshoot and fix bib formatting. Bib file works now, but some citations recieve double brackets. Will continue looking into this issue
csl: elsevier.csl
---

This document will be my final manuscript for the MADA project. I am keeping these draft sections for structure now, until I edit them later.

# Summary/Abstract
_Write a summary of your project._

@@

# Methods and Results

_In most research papers, results and methods are separate. You can combine them here if you find it easier. You are also welcome to structure things such that those are separate sections._

@@

>>>>

---
title: "Plastic in the Urban Freshwater Environment"
subtitle: "An Exploratory Study of Microplastics in the Upper Oconee Watershed via Community-Based Efforts"
author: Amelia M. Foley
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::word_document2: 
    toc: false
    number_sections: true
    #reference_docx: style_template.docx #working on creating a template doc with the appropriate style for the conference I am attending next summer (ASTM) so that my manuscript can be automatically formatted
  bookdown::html_document2: 
    toc: false
bibliography: library.bib 
csl: elsevier.csl
---

# Abstract
With growing awareness of plastic pollution and its detrimental environmental impacts, there has been increased attention on the study of microplastics. These small debris result from the weathering and breakdown of plastic waste from urban runoff, agricultural runoff, ocean dumping, stormwater, waste management, and other sources. Microplastics pose a threat to both aquatic life and human health due to their chemical makeup, their build-up in the environment, and their suspected long-term biological effects. As their ubiquity becomes more apparent, so does the need for accessible, affordable, and effective methods for the widespread study and quantification of microplastic pollution.

This study takes place in Athens, GA, an urban area which houses a population of about 127,000. The project relies on local volunteer citizen science efforts for water sampling, processing, and analysis. Given the local and regional ecological significance of the Oconee River—including its transport of freshwater to the Atlantic Ocean—this study aims to collect relevant data regarding microplastics in the Upper Oconee Watershed, which includes 2 rivers and 15 creeks in Athens-Clarke County.

Through community-based sample collection and analysis, volunteers quantified microplastics through membrane filtration and visual identification. Preliminary data consistently show that local freshwater microplastic levels are greater than the U.S. average in the Upper Oconee Watershed, a portion of the Oconee River basin with both a high population size and high population density relative to other regions of Georgia. Average and median concentrations of 95 particles/L and 46 particles/L, respectively, were recorded throughout one year of seasonal sampling. Further visual and statistical analysis will confirm the project’s results, and discussion of these results in context will reveal the efficacy of this community-based approach to the quantification of microplastics.

@@

# Methods

@@

# Results
